Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-09 agujournaltemplate.tex
editing May 9th

\title{Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela}

@@

\authors{B. Post\affil{1,2}, E. Acevedo-Trejos\affil{3}, S. Chakraborty\affil{1}, A.D. Barton\affil{4}, A. Merico\affil{1}}

@@

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

@@

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

@@

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots.}
\label{fig:clustering}
\end{figure}

@@

This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}. These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO". 
We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased uwpelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}). At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

@@

However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}. We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that actually separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2. This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1. Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.

@@

The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag. We also tested a 12 month lag and found no significant improvement.
The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions. For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events. Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range. Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).

>>>>

\title{Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela}

@@

\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1}}

@@

%Comments from Subhi:
- last 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
- streamline discussion, much shorter!
-

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

@@

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% ADD Duplicate Points here:
    During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

    Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".  
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

    %% Gaps that We Adresse:
    % - More in-depth exploration of microscopy data & diversity time series
    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."

@@

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$). 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

% ADD HERE MORE DISCUSSION OF MODEL COMPARISON

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
\label{fig:clustering}
\end{figure}

@@

This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased upwelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}). At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

@@

However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.

The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}. We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

@@

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2. This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1. Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.

@@

% Move the general point on ENSO to intro! Only Discussion-relevant part here..
The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag. We also tested a 12 month lag and found no significant improvement.
The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions. For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events. Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range. Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).

The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, temperature and salinity, in that order. Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}. For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups. Other functional groups including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance by very strong upwelling. Similarly, for temperature, dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling. For the other predictors, we do not see strong threshold responses.
